Practical Optional Types for Clojure

### 2015-02-23 formal-model.tex
'update' covers all cases for ClassPE

Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \mintinline{clojure}{:a}), 
constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.

@@

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \mintinline{clojure}{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \mintinline{clojure}{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \mintinline{clojure}{(inc :a)} is unreachable (and it clearly is by BE-Do1).

This works similarly with our motivating example. Intuitively, when evaluating
\mintinline{clojure}{(inc x)} we know \mintinline{clojure}{(assert (number? x))} has evaluated to a value,
so we assume \mintinline{clojure}{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \mintinline{clojure}{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \mintinline{clojure}{x} as a \Number.

@@

B-IsA with the metafunction \isaopsemliteral{} give the operational semantics.

\smallskip
$
\begin{array}{ll}
  \vcenter{\hbox{\BIsA{}}}
  &
  \vcenter{\hbox{\isaopsemfigure{}}}
\end{array}
$
\smallskip

For the typing rules, we add rules for type checking keyword and class literals
as singleton types, and define T-IsA, which infers the propositions of an \isaliteral{} 
expression by passing
 the \emph{object} of the left argument and the \emph{type}
of the right argument to \isacompareliteral{}.

\smallskip
$
\begin{array}{ll}
\vcenter{\hbox{
    \TIsA{}}} &

  \vcenter{\hbox{$
    \begin{array}{l}
      {\TKw}\\\\
      {\TClass}
    \end{array}$}}
\end{array}
$
\smallskip

\begin{figure*}
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}
  \caption{\isaopsemliteral Proposition set Metafunction}
\end{figure*}

\begin{mathpar}
    {\TKw}

    {\TClass}

    \isapropsfigure{}

    \TIsA{}
\end{mathpar}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isapropsfigure{}

    \TIsA{}
  \end{mathpar}
  \caption{\isaliteral{} typing rules}
  \label{main:figure:isatyping}
\end{figure}

@@

A multimethod in Clojure is a function with a dispatch mechanism to
of internal methods. Multimethods are created with {\mintinline{clojure}{defmulti}}
and methods are installed via {\mintinline{clojure}{defmethod}}.
They are called like regular functions.

@@

%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\mintinline{clojure}{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%                       {\classconst{}}}
%                     {\\ bar}$

>>>>

Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \clj{:a}), 
constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.

@@

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \clj{(inc :a)} is unreachable (and it clearly is by BE-Do1).

This works similarly with our motivating example. Intuitively, when evaluating
\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
so we assume \clj{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \clj{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \clj{x} as a \Number.

@@

B-IsA with the metafunction \isaopsemliteral{} reflect these semantics.

$$
\begin{array}{ll}
  \vcenter{\hbox{\BIsA{}}}
  &
  \vcenter{\hbox{\isaopsemfigure{}}}
\end{array}
$$

For the typing rules, we add T-Kw and T-Class for type checking keyword and class literals
as singleton types, and T-IsA for \isaliteral{} expressions, which 
uses the \emph{object} of the left argument and the \emph{type} of the right argument
to infer the final proposition set.

$$
\begin{array}{ll}
\vcenter{\hbox{
    \TIsA{}}} &

  \vcenter{\hbox{$
    \begin{array}{l}
      {\TKw}\\\\
      {\TClass}
    \end{array}$}}
\end{array}
$$

The definition of \isacompareliteral{} (figure~\ref{main:figure:isaprops}) is deliberately conservative.
The first line considers the case where the object of the left argument
is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
The simplest example

\begin{minted}{clojure}
  (isa? (class x) Keyword)
\end{minted}

has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
that is if the expression is true then \clj{x} is a keyword, otherwise \clj{x} is not a keyword.

We can also finally define \nil{} test

\begin{minted}{clojure}
(ann nil? (Pred nil))
(defn nil? [x]
  (isa? (class (class x)) Class))
\end{minted}

with the proposition set 
\isacompare{\s{}}{\classpe{}({\path{\classpe{}}{\x{}}})}{\Value{\Class}}{\filterset{\isprop{\Class}{\path{\classpe{}}{\x{}}}}{\notprop{\Class}{\path{\classpe{}}{\x{}}}}},

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}

  \center{Takes \emph{object} of the left parameter and \emph{type} of the right parameter.}
  \caption{Proposition Set for \isaliteral{} expressions}
  \label{main:figure:isaprops}
\end{figure}

@@

@@

A multimethod in Clojure is a function with a dispatch mechanism to
of internal methods. Multimethods are created with {\clj{defmulti}}
and methods are installed via {\clj{defmethod}}.
They are called like regular functions.

@@

%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\clj{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%                       {\classconst{}}}
%                     {\\ bar}$
